BIP-Predictor

### 2025-05-21 Sweave/BIP_predictor.Rnw
up

%As shown in Wildi (2025), the SSA criterion's objective function—the target correlation $\rho(y,z,h)$—can be linked to the sign accuracy of the predictor $SA(y_t):=\text{P}\left(\text{sign}(z_{t+h}) = \text{sign}(y_t)\right)$, i.e., the probability that the signs of the predictor $y_t$ and the target $z_{t+h}$ agree:
%\begin{eqnarray}\label{sig_acc}
%SA(y_t)=0.5+\frac{\arcsin(\rho(y,z,h))}{\pi}.
%\end{eqnarray}
%Due to the strict monotonicity of the arcsin function, maximizing the target correlation also maximizes the sign accuracy of the predictor.\footnote{Equation \ref{sig_acc} holds exactly for Gaussian processes. Due to the central limit theorem, the relationship holds approximately for low-pass filtered non-Gaussian series, see Wildi (2025).} 
\begin{itemize}
\item Alternative to MSE, target correlation, holding-time: hit-rate, false alarm. 
\item We can link objective function (target correlation) with sign accuracy, see Wildi (2025). BUT: risk that reviewer asks for comparison with logit or whatever sign-forecast tool.
\item Context ROC-curve
\end{itemize}
For the ROC curves, we consider predictors of GDP based on both original (unfiltered) and filtered indicators. For the direct forecast, we use the regression
\[
GDP_{t+shift}=\beta_0^{shift}+\sum_{i=1}^5\beta_i^{shift}x_{ti}+\epsilon_{t}^{shift,direct}=\hat{GDP}_{t+shift}^{direct}+\epsilon_{t}^{shift,direct},
\]
where $x_{ti}$ designates the original indicators. 
For the filtered indicators, we consider similar regression equations:
\[GDP_{t+shift}=\beta_0^{shift,h,filt}+\sum_{i=1}^5\beta_i^{shift,h,filt}y_{ti}^{h,filt}+\epsilon_{t}^{h,shift,filt}=\hat{GDP}_{t+shift}^{filt}+\epsilon_{t}^{shift,filt},\]
where $y_{ti}^{h,filt}$ represents the $i$-th filtered indicator, based on either HP-C, M-MSE, or M-SSA. For simplicity of exposition, we present results based on the full sample, excluding the pandemic period. Fig. 
<<label=init,results=hide>>=

@@

# Combination of forward-shift (of target) and forecast horizon (underlying optimization)
shift_vec<-1:5
hh_vec<-1:6
AUC_array_GDP<-AUC_array_HP_GDP<-array(dim=c(length(shift_vec),length(hh_vec),4))
dimnames(AUC_array_GDP)<-dimnames(AUC_array_HP_GDP)<-list(paste("shift=",shift_vec,sep=""),paste("h=",hh_vec,sep=""),c("Direct forecast","Direct HP forecast","M-MSE","M-SSA"))
# Select shifts>k (otherwise too many plots in single window)
k<-2
# Show plot
showl<-T
# Labels: hit rate vs. false alarm rate
lbls = "Hit"

@@

#file = "./Figures/ROC_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))

@@

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
    AUC_array_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

@@

#file = "./Figures/ROC_HP_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))

@@

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
    AUC_array_HP_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

@@

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-T
AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
shift<-4
h<-shift

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-F
AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)

@@

mat1<-round(auc_GDP_mat,3)
#latex(cor_vec, dec = 1, , caption = "Example of using latex to create table",
#center = "centering", file = "", floating = FALSE)
xtable(mat1, dec = 1,digits=rep(3,dim(mat1)[2]+1),
paste("AUCs of different predictors against GDP at various forecast horizons.."),
label=paste("p_val_wc5",sep=""),
center = "centering", file = "", floating = FALSE)
@

>>>>

%As shown in Wildi (2025), the SSA criterion's objective function—the target correlation $\rho(y,z,h)$—can be linked to the sign accuracy of the predictor $SA(y_t):=\text{P}\left(\text{sign}(z_{t+h}) = \text{sign}(y_t)\right)$, i.e., the probability that the signs of the predictor $y_t$ and the target $z_{t+h}$ agree:
%\begin{eqnarray}\label{sig_acc}
%SA(y_t)=0.5+\frac{\arcsin(\rho(y,z,h))}{\pi}.
%\end{eqnarray}
%Due to the strict monotonicity of the arcsin function, maximizing the target correlation also maximizes the sign accuracy of the predictor.\footnote{Equation \ref{sig_acc} holds exactly for Gaussian processes. Due to the central limit theorem, the relationship holds approximately for low-pass filtered non-Gaussian series, see Wildi (2025).} 
\textbf{Remarks:}
\begin{itemize}
\item Alternative to MSE, target correlation, holding-time: hit-rate, false alarm. 
\item We can link objective function (target correlation) with sign accuracy, see Wildi (2025). BUT: risk that reviewer asks for comparison with logit or whatever sign-forecast tool.
\item Context ROC-curve
\end{itemize}
For the ROC curves, we consider predictors of GDP based on both original (unfiltered) and filtered indicators. For the direct forecast, we use the regression
\[
GDP_{t+shift}=\beta_0^{shift}+\sum_{i=1}^5\beta_i^{shift}x_{ti}+\epsilon_{t}^{shift,direct}=\hat{GDP}_{t+shift}^{direct}+\epsilon_{t}^{shift,direct},
\]
where $x_{ti}$ designates the original indicators. 
For the filtered indicators, we consider similar regression equations:
\[GDP_{t+shift}=\beta_0^{shift,h,filt}+\sum_{i=1}^5\beta_i^{shift,h,filt}y_{ti}^{h,filt}+\epsilon_{t}^{h,shift,filt}=\hat{GDP}_{t+shift}^{h,filt}+\epsilon_{t}^{h,shift,filt},\]
where $y_{ti}^{h,filt}$ represents the $i$-th filtered indicator optimized at forecast horizon $h$, based on either HP-C, M-MSE, or M-SSA. \\
\textbf{Remarks:}
\begin{itemize}
\item I'm computing all combinations of shift and h. 
\item But we could emphasize the diagonal only (h=shift): simplifies notation/explanation.
\item The table of AUCs (below) is based on diagonal only
\end{itemize}
For simplicity of exposition, we present results based on the full sample, excluding the pandemic period. Fig.\ref{ROC_GDP_shift_1_4} compares ROCs of the predictors for small and mid-term forecast horizons and Table \ref{auc} presents AUCs (area under the curve) for forecast horizons 1-5 quarters ahead.
<<label=init,results=hide>>=

@@

# Combination of forward-shift (of target) and forecast horizon (underlying optimization)
shift_vec<-1:5
hh_vec<-1:6
AUC_array_GDP<-AUC_array_HP_GDP<-array(dim=c(length(shift_vec),length(hh_vec),4))
dimnames(AUC_array_GDP)<-dimnames(AUC_array_HP_GDP)<-list(paste("shift=",shift_vec,sep=""),paste("h=",hh_vec,sep=""),c("Direct forecast","Direct HP forecast","M-MSE","M-SSA"))
# Select shifts>k (otherwise too many plots in single window)
k<-2
# Labels: hit rate vs. false alarm rate
lbls = "Hit"
# Size of legend box in plot
lg_cex<-0.5

@@

#file = "./Figures/ROC_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))
# Show legend
showl<-T

@@

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = lg_cex, lg_ncol = 1)
    AUC_array_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

@@

#file = "./Figures/ROC_HP_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))
showl<-T
for (i in 1:length(shift_vec))#i<-1
{ 
  for (j in 1:length(hh_vec))#j<-1
  {
    shift<-shift_vec[i]
    h<-hh_vec[j]

@@

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = lg_cex, lg_ncol = 1)
    AUC_array_HP_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

@@

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-T
AUC<-ROCplots(ROC_data, showROC , main = "One quarter ahead", lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
shift<-4
h<-shift

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-F
AUC<-ROCplots(ROC_data, showROC , main = "One year ahead", lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)

@@

mat1<-round(auc_GDP_mat,3)
#latex(cor_vec, dec = 1, , caption = "Example of using latex to create table",
#center = "centering", file = "", floating = FALSE)
xtable(mat1, dec = 1,digits=rep(3,dim(mat1)[2]+1),
paste("AUCs of predictors against GDP at various forecast horizons."),
label=paste("p_val_wc5",sep=""),
center = "centering", file = "", floating = FALSE)
@
At forecast horizons less than or equal to two quarters, the direct forecast challenges the filters. However, for horizons $h>2$, the filters outperform the direct approach: the univariate HP-C design is surpassed by the classic multivariate MSE filter, which is in turn dominated by the M-SSA. These results confirm that controlling the rate of sign changes of the predictor, while maximizing its target correlation through M-SSA can enhance predictor performance in terms of sign accuracy and false alarm rate.
